---
title: "Developing KDI pipeline"
author: "Abel Torres Espin"
date: '2025-04-03'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
library(tidyverse)
library(DT)
library(readxl)
library(forecast) ## for time series functionalities
library(lme4)
library(lmerTest)
library(gganimate)
library(splines)
library(vegan)
library(patchwork)
```

## Utility functions

### Plotting functions

The following function allows for the plotting of all the data of an animal and reach type

```{r}
plot_reach<-function(df, reach.type = "successful"){
  df<-df%>%
    filter(reach_type ==reach.type)
  
  start<-df%>%
    group_by(reach)%>%
    mutate(start = ifelse(frame == 1, "start", NA))%>%
    filter(!is.na(start))

  end<-df%>%
    group_by(reach)%>%
    mutate(end = ifelse(frame == max(frame, na.rm = T), "end", NA))%>%
    filter(!is.na(end))

  pellet_center_df<-df%>%
    group_by(reach)%>%
    filter(frame %in% 1:10)%>%
    summarise(x_p_smooth = median(x_p_smooth), y_p_smooth = median(y_p_smooth))

  plot<-df%>%
    ggplot(aes(x_smooth, y_smooth))+
    geom_path(alpha = 1, show.legend = F, aes(group = reach))+
    geom_point(data = start, color = "red")+
      geom_point(data = end, color = "green")+
      geom_point(data = pellet_center_df, aes(x = x_p_smooth, y = y_p_smooth),color = "purple")+
      geom_path(data = df, aes(x = x_p_smooth, y = y_p_smooth), color = "purple")+
    facet_grid(digit~reach)+
    theme_bw()+
    ggtitle(paste(reach.type, unique(df$animal_id)))
  
  return(plot)
}
```

### KDI functions

#### GPA function
Function to run GPA as for Gower algorithm through the 2 step process.

```{r}
## GPA function
GPA<-function(X_list, tolerance = 0.0001){
  
  m<-length(X_list)
  
  n_row <-nrow(X_list[[1]])
  n_col <-ncol(X_list[[1]])
  
  ## initialize set list of T transformation from m = 2 to m = m with identity
  T0<-diag(nrow = n_col,ncol = n_col)
  
  ## initialize result list
  AT_list<-lapply(X_list, function(x){
    as.matrix(x) %*% T0
  })
  
  ## Internal function for the iteration of steps 2 to M
  proc_2M_steps<-function(X_list, m, AT_list){
    for (i in 1:m){
      if (i == 1){
        B<-apply(simplify2array(AT_list[2:m]), 1:2, sum)
      }else if (i %in% 2:m){
        AT_fit <- apply(simplify2array(AT_list[1:(i-1)]), 1:2, sum)
        B<-AT_fit+apply(simplify2array(AT_list[i:m]), 1:2, sum)
      }else if (i == m){
        B<-apply(simplify2array(AT_list[1:(m-1)]), 1:2, sum)
      }
      fit<-pracma::procrustes(B,AT_list[[i]])
      AT_list[[i]]<-fit$P## the matrix P is AT
    }
    
    return(list("AT_list" = AT_list))
  }
  
  ## sequence of 10
  res_AT_list<-list()
  Sr<-list()
  Y<-list()
  
  ## first sequence
  first_2M<-proc_2M_steps(X_list, m = m, AT_list)
  res_AT_list[[1]]<-first_2M$AT_list
  Y[[1]]<-apply(simplify2array(res_AT_list[[1]]), 1:2, mean)
  Sr[[1]]<-sum(diag((Y[[1]]%*%t(Y[[1]]))))
  
  
  Sr_diff<-1 
  t<-2 ## iteration counter for convergence
  
  while(Sr_diff>tolerance){
    
    t_2M<-proc_2M_steps(res_AT_list[[t-1]],m=m, res_AT_list[[t-1]])
    res_AT_list[[t]]<-t_2M$AT_list
    
    Y[[t]]<-apply(simplify2array(res_AT_list[[t]]), 1:2, mean)
    Sr[[t]]<-sum(diag((Y[[t]]%*%t(Y[[t]]))))
    
    Sr_diff<-abs(Sr[[t]]-Sr[[t-1]])
    t<-t+1
  }
  
  results<-list("consensus" = Y[[t-1]], "iterations" = t-1,
                "fitted_X" = res_AT_list[[t-1]])
  
  return(results)
}
```

# Cohort 1: NT study
## Read data

This section reads the files and process them to have them in a single dataset for downstream processing.

### Pre injury
```{r}
unsuccess_data<-read_xlsx("data/original/Uninjured pca data all digits.xlsx", sheet = 1)
unsuccess_data$reachtype<-"unsuccessful"

success_data<-read_xlsx("data/original/Uninjured pca data all digits.xlsx", sheet = 2)
success_data$reachtype<-"successful"

attempt_pellet_data<-read_xlsx("data/original/Uninjured pca data all digits.xlsx", sheet = 3)
attempt_pellet_data$reachtype<-"attempt_pellet"

attempt_Nopellet_data<-read_xlsx("data/original/Uninjured pca data all digits.xlsx", sheet = 4)
attempt_Nopellet_data$reachtype<-"attempt_Nopellet"

raw_data<-rbind(unsuccess_data, success_data, attempt_pellet_data, attempt_Nopellet_data)

colnames(raw_data)<-c("frame", "x_d1", "x_d2", "x_d3", "x_d4", "y_d1", "y_d2", 
                       "y_d3", "y_d4", "x_p", "y_p","digit_spread", "pellet_d1",
                       "pellet_d2", "pellet_d3", "pellet_d4",
                       "angle", "animal_id", "reach_type") 

raw_data<-raw_data%>%
  arrange(animal_id)

raw_data$reach<-NA
count_reaches<-0
animal<-"4393"
for (i in 1:nrow(raw_data)){
  if (raw_data$frame[i] == 1 & raw_data$animal_id[i] == animal){
    count_reaches<-count_reaches+1
  }else if (raw_data$animal_id[i] != animal){
    animal<-raw_data$animal_id[i]
    count_reaches<-1
  }
  raw_data$reach[i]<-count_reaches
}

raw_data$timepoint<-"pre_injury"
```

### Post SCI
```{r}
unsuccess_data_post<-read_xlsx("data/original/Post-SCI pca data all digits.xlsx", sheet = 1)
unsuccess_data_post$reachtype<-"unsuccessful"

success_data_post<-read_xlsx("data/original/Post-SCI pca data all digits.xlsx", sheet = 2)
success_data_post$reachtype<-"successful"

attempt_pellet_data_post<-read_xlsx("data/original/Post-SCI pca data all digits.xlsx", sheet = 3)
attempt_pellet_data_post$reachtype<-"attempt_pellet"

attempt_Nopellet_data_post<-read_xlsx("data/original/Post-SCI pca data all digits.xlsx", sheet = 4)
attempt_Nopellet_data_post$reachtype<-"attempt_Nopellet"

raw_data_post<-rbind(unsuccess_data_post, success_data_post, attempt_pellet_data_post, attempt_Nopellet_data_post)

colnames(raw_data_post)<-c("frame", "x_d1", "x_d2", "x_d3", "x_d4", "y_d1", "y_d2", 
                       "y_d3", "y_d4", "x_p", "y_p","digit_spread", "pellet_d1",
                       "pellet_d2", "pellet_d3", "pellet_d4",
                       "angle", "animal_id", "reach_type") 

raw_data_post<-raw_data_post%>%
  arrange(animal_id)

raw_data_post$reach<-NA
count_reaches<-0
animal<-"4393"
for (i in 1:nrow(raw_data_post)){
  if (raw_data_post$frame[i] == 1 & raw_data_post$animal_id[i] == animal){
    count_reaches<-count_reaches+1
  }else if (raw_data_post$animal_id[i] != animal){
    animal<-raw_data_post$animal_id[i]
    count_reaches<-1
  }
  raw_data_post$reach[i]<-count_reaches
}

raw_data_post$timepoint<-"post-SCI"
```

### 8Weeks SCI
```{r}
unsuccess_data_8w<-read_xlsx("data/original/Wk8 pca data all digits.xlsx", sheet = 1)
unsuccess_data_8w$reachtype<-"unsuccessful"

# success_data_8w<-read_xlsx("data/original/Wk8 pca data all digits.xlsx", sheet = 2)
# success_data_8w$reachtype<-"successful"

attempt_pellet_data_8w<-read_xlsx("data/original/Wk8 pca data all digits.xlsx", sheet = 2)
attempt_pellet_data_8w$reachtype<-"attempt_pellet"

attempt_Nopellet_data_8w<-read_xlsx("data/original/Wk8 pca data all digits.xlsx", sheet = 3)
attempt_Nopellet_data_8w$reachtype<-"attempt_Nopellet"

raw_data_8w<-rbind(unsuccess_data_8w,attempt_pellet_data_8w, attempt_Nopellet_data_8w)

colnames(raw_data_8w)<-c("frame", "x_d1", "x_d2", "x_d3", "x_d4", "y_d1", "y_d2", 
                       "y_d3", "y_d4", "x_p", "y_p","digit_spread", "pellet_d1",
                       "pellet_d2", "pellet_d3", "pellet_d4",
                       "angle", "animal_id", "reach_type") 

raw_data_8w<-raw_data_8w%>%
  arrange(animal_id)

raw_data_8w$reach<-NA
count_reaches<-0
animal<-"4393"
for (i in 1:nrow(raw_data_8w)){
  if (raw_data_8w$frame[i] == 1 & raw_data_8w$animal_id[i] == animal){
    count_reaches<-count_reaches+1
  }else if (raw_data_8w$animal_id[i] != animal){
    animal<-raw_data_8w$animal_id[i]
    count_reaches<-1
  }
  raw_data_8w$reach[i]<-count_reaches
}

raw_data_8w$timepoint<-"8w-SCI"
```

### 16 Weeks SCI
```{r}
unsuccess_data_16w<-read_xlsx("data/original/Wk16 pca data all digits.xlsx", sheet = 1)
unsuccess_data_16w$reachtype<-"unsuccessful"

# success_data_16w<-read_xlsx("data/original/Wk16 pca data all digits.xlsx", sheet = 2)
# success_data_16w$reachtype<-"successful"

attempt_pellet_data_16w<-read_xlsx("data/original/Wk16 pca data all digits.xlsx", sheet = 2)
attempt_pellet_data_16w$reachtype<-"attempt_pellet"

attempt_Nopellet_data_16w<-read_xlsx("data/original/Wk16 pca data all digits.xlsx", sheet = 3)
attempt_Nopellet_data_16w$reachtype<-"attempt_Nopellet"

raw_data_16w<-rbind(unsuccess_data_16w,attempt_pellet_data_16w, attempt_Nopellet_data_16w)

colnames(raw_data_16w)<-c("frame", "x_d1", "x_d2", "x_d3", "x_d4", "y_d1", "y_d2", 
                       "y_d3", "y_d4", "x_p", "y_p","digit_spread", "pellet_d1",
                       "pellet_d2", "pellet_d3", "pellet_d4",
                       "angle", "animal_id", "reach_type") 

raw_data_16w<-raw_data_16w%>%
  arrange(animal_id)

raw_data_16w$reach<-NA
count_reaches<-0
animal<-"4393"
for (i in 1:nrow(raw_data_16w)){
  if (raw_data_16w$frame[i] == 1 & raw_data_16w$animal_id[i] == animal){
    count_reaches<-count_reaches+1
  }else if (raw_data_16w$animal_id[i] != animal){
    animal<-raw_data_16w$animal_id[i]
    count_reaches<-1
  }
  raw_data_16w$reach[i]<-count_reaches
}

raw_data_16w$timepoint<-"16w-SCI"
```

### Join datasets

```{r}
raw_data_all<-bind_rows(raw_data, raw_data_post, raw_data_8w, raw_data_16w)
```

## Clean all data

```{r}
# create unique reach identifier
raw_data_all<-raw_data_all%>%
  mutate(reach_id = paste(timepoint, animal_id, reach, sep = "_"))

raw_data_all_long<-raw_data_all%>%
  pivot_longer(cols = c(x_d1, x_d2, x_d3, x_d4, y_d1, y_d2, y_d3, y_d4, 
                        pellet_d1, pellet_d2, pellet_d3, pellet_d4))%>%
  mutate(digit=str_split(name, "_", simplify = T)[,2], 
         pos_var = str_split(name, "_", simplify = T)[,1],
         pos_var = ifelse(pos_var == "pellet", "dist_pellet", pos_var))%>%
  select(-name)%>%
  pivot_wider(id_cols = c(animal_id, timepoint, reach_id, frame,digit, x_p, y_p,
                          digit_spread, angle, reach_type), names_from = pos_var)

## Calculate distances between digits
raw_data_all<-raw_data_all%>%
  group_by(reach_id, frame)%>%
  mutate(d1_d2 = dist(rbind(c(x_d1, y_d1),
                            c(x_d2, y_d2))),
         d2_d3 = dist(rbind(c(x_d2, y_d2),
                            c(x_d3, y_d3))),
         d3_d4 = dist(rbind(c(x_d3, y_d3),
                            c(x_d4, y_d4))))

# save dataset for future use
saveRDS(raw_data_all, "raw_data_all.Rda")
```

```{r}
#load raw all dataset
raw_data_all<-readRDS("raw_data_all.Rda")

#determine quantile thresholds
thr_d1_d2<-quantile(raw_data_all$d1_d2, probs = 0.92)
thr_d2_d3<-quantile(raw_data_all$d2_d3, probs = 0.92)
thr_d3_d4<-quantile(raw_data_all$d3_d4, probs = 0.92)

#Eliminate extremes filtered based on the distance between digits
clean_data_all<-raw_data_all%>%
  ungroup()%>%
  rowwise()%>%
  mutate(x_d1 = ifelse(d1_d2>thr_d1_d2, NA, x_d1),
         y_d1 = ifelse(d1_d2>thr_d1_d2, NA, y_d1),
         x_d2 = ifelse(d2_d3>thr_d2_d3 | d1_d2>thr_d1_d2, NA, x_d2),
         y_d2 = ifelse(d2_d3>thr_d2_d3 | d1_d2>thr_d1_d2, NA, y_d2),
         x_d3 = ifelse(d2_d3>thr_d2_d3 | d3_d4>thr_d3_d4, NA, x_d3),
         y_d3 = ifelse(d2_d3>thr_d2_d3 | d3_d4>thr_d3_d4, NA, y_d3),
         x_d4 = ifelse(d3_d4>thr_d3_d4, NA, x_d4),
         y_d4 = ifelse(d3_d4>thr_d3_d4, NA, y_d4))

# Calculate mahalanobis distance
out<-psych::outlier(clean_data_all[,c(2,9)])

clean_data_all<-clean_data_all%>%
  ungroup()%>%
  mutate(out = out,
         filter_out = out>10)%>%## filter based on mahalanobis distance
  filter(filter_out == FALSE)
```


```{r}
# Summarise based on proportion of NA values. Those reaches with high NA (more tha 20%) are excluded
clean_summary<-clean_data_all%>%
  group_by(reach_id)%>%
  summarise_at(.vars = c("x_d1", "y_d1", "x_d2", "y_d2", "x_d3", "y_d3", "x_d4", "y_d4"),
               .funs = function(x) {mean(is.na(x))})

clean_summary$means_nas<-rowMeans(clean_summary[,-1])
reach_to_keep<-clean_summary$reach_id[which(clean_summary$means_nas<=0.2)]

clean_data_all<-clean_data_all%>%
  filter(reach_id%in%reach_to_keep)

# Transform to long format
clean_data_all_long<-clean_data_all%>%
  pivot_longer(cols = c(x_d1, x_d2, x_d3, x_d4, y_d1, y_d2, y_d3, y_d4, 
                        pellet_d1, pellet_d2, pellet_d3, pellet_d4))%>%
  mutate(digit=str_split(name, "_", simplify = T)[,2], 
         pos_var = str_split(name, "_", simplify = T)[,1],
         pos_var = ifelse(pos_var == "pellet", "dist_pellet", pos_var))%>%
  select(-name)%>%
  pivot_wider(id_cols = c(animal_id, timepoint, reach_id, frame,digit, x_p, y_p,
                          digit_spread, angle, reach_type), names_from = pos_var)


## Use tsclean and supsmu to clean temporal outliers and smooth
clean_data_all_s<-clean_data_all_long%>%
  group_by(reach_id, digit)%>%
  mutate(n_frame= n())%>%
  filter(n_frame>=20)%>%
  filter(frame<=(n_frame-5))%>%
  mutate(clean_x = tsclean(x),
         clean_y = tsclean(y),
         clean_x_p = tsclean(x_p),
         clean_y_p = tsclean(y_p))%>%
  mutate(smooth_x = supsmu(x = frame, y = clean_x, span = 0.05)$y,
         smooth_y = supsmu(x = frame, y = clean_y, span = 0.05)$y,
         smooth_x_p = supsmu(x = frame, y = clean_x_p, span = 0.05)$y,
         smooth_y_p = supsmu(x = frame, y = clean_y_p, span = 0.05)$y)

## Add features
clean_data_all_wide<-clean_data_all_s%>%
  pivot_wider(id_cols = c(animal_id, timepoint, reach_id, frame, reach_type),
              names_from = "digit",
                values_from = c("smooth_x","smooth_y", "x", "y"))

clean_data_all_wide<-clean_data_all_wide%>%
  group_by(reach_id, frame)%>%
  mutate(d1_d4 = as.numeric(dist(rbind(c(smooth_x_d1, smooth_y_d1),
                            c(smooth_x_d4, smooth_y_d4)))),
         angle = sin((smooth_x_d1-smooth_x_d4)/d1_d4)*57.29)%>%
  group_by(reach_id)%>%
  mutate(d1_x_speed = abs(c(0,diff(smooth_x_d1))),
         d1_y_speed = abs(c(0,diff(smooth_y_d1))),
         d2_x_speed = abs(c(0,diff(smooth_x_d2))),
         d2_y_speed = abs(c(0,diff(smooth_y_d2))),
         d3_x_speed = abs(c(0,diff(smooth_x_d3))),
         d3_y_speed = abs(c(0,diff(smooth_y_d3))),
         d4_x_speed = abs(c(0,diff(smooth_x_d4))),
         d4_y_speed = abs(c(0,diff(smooth_y_d4))))
         

clean_data_all_wide%>%
    filter(reach_id == "pre_injury_4873_20")%>%
  ggplot()+
  geom_segment(aes(x = smooth_x_d1, xend = smooth_x_d4, 
                   y = smooth_y_d1, yend = smooth_y_d4), color = "grey")+
    geom_point(aes(smooth_x_d1, smooth_y_d1))+
  geom_point(aes(smooth_x_d4, smooth_y_d4), color = "red")+
  # geom_text(aes(smooth_x_d4, smooth_y_d4, label = round(angle,2)))+
    # geom_line(aes(frame, b), color = "red")+
  facet_wrap(~frame)

clean_data_all_wide%>%
    filter(animal_id == "4873", reach_type == "successful")%>%
  ggplot()+
  geom_line(aes(frame, d4_y_speed))+
    # geom_line(aes(frame, b), color = "red")+
  facet_wrap(~reach_id)

saveRDS(clean_data_all_wide, "clean_data_all_wide.Rds")
```

### Check cleaning

```{r}
raw_data_all%>%
    filter(timepoint == "pre_injury", reach_type == "successful",
         animal_id == "4875")%>%
  ggplot(aes(x_d3,y_d4))+
  geom_path()+
  geom_path(data = clean_data_all_wide%>%
    filter(timepoint == "pre_injury", reach_type == "successful",
         animal_id == "4875"), aes(x_d3, y_d3), color = "red")+
  facet_wrap(~reach_id)

raw_data_all%>%
    filter(timepoint == "pre_injury", reach_type == "successful",
         animal_id == "4875")%>%
  ggplot(aes(frame,y_d3))+
  geom_path()+
  geom_path(data = clean_data_all_wide%>%
    filter(timepoint == "pre_injury", reach_type == "successful",
         animal_id == "4875"), aes(frame, y_d3), color = "red")+
  facet_wrap(~reach_id)

clean_data_all_wide_norm%>%
  filter(animal_id=="4875", reach_type == "successful",timepoint == "pre_injury")%>%
  ggplot(aes(smooth_x_d1, smooth_y_d1, group = reach_id))+
  geom_path()+
  geom_path(aes(x=smooth_x_d2, y = smooth_y_d2), color = "red")+
  geom_path(aes(x=smooth_x_d3, y = smooth_y_d3), color = "orange")+
  geom_path(aes(x=smooth_x_d4, y = smooth_y_d4), color = "blue")+
  facet_wrap(~reach_id)

raw_data_all%>%
    filter(animal_id=="4904", reach_type == "successful", reach == 41)%>%
  ggplot(aes(x_d1, y_d1))+
  geom_path()+
  geom_path(aes( x = x_d2, y = y_d2), color = "red")+
  geom_path(aes( x = x_d3, y = y_d3), color = "orange")+
  geom_path(aes( x = x_d4, y = y_d4), color = "blue")

raw_data_all_long%>%
    filter(animal_id=="4904", reach_type == "successful")%>%
  ggplot(aes(frame, x, color = digit))+
  geom_path()+
  facet_wrap(~reach_id)
```

### time normalization
```{r}
clean_data_all_wide_norm<-clean_data_all_wide%>%
  group_by(reach_id)%>%
  summarise(animal_id = unique(animal_id),
            frame = 1:100,
            timepoint = unique(timepoint),
            reach_type = unique(reach_type),
            smooth_x_d1 = approx(x = 1:n(), y = smooth_x_d1, n = 100)$y,
            smooth_x_d2 = approx(x = 1:n(), y = smooth_x_d2, n = 100)$y, 
            smooth_x_d3 = approx(x = 1:n(), y = smooth_x_d3, n = 100)$y, 
            smooth_x_d4 = approx(x = 1:n(), y = smooth_x_d4, n = 100)$y, 
            smooth_y_d1 = approx(x = 1:n(), y = smooth_y_d1, n = 100)$y, 
            smooth_y_d2 = approx(x = 1:n(), y = smooth_y_d2, n = 100)$y, 
            smooth_y_d3 = approx(x = 1:n(), y = smooth_y_d3, n = 100)$y, 
            smooth_y_d4 = approx(x = 1:n(), y = smooth_y_d4, n = 100)$y,
            d1_d4 = approx(x = 1:n(), y = d1_d4, n = 100)$y,
            angle = approx(x = 1:n(), y = angle, n = 100)$y,
            d1_x_speed = approx(x = 1:n(), y = d1_x_speed, n = 100)$y,
            d2_x_speed = approx(x = 1:n(), y = d2_x_speed, n = 100)$y,
            d3_x_speed = approx(x = 1:n(), y = d3_x_speed, n = 100)$y,
            d4_x_speed = approx(x = 1:n(), y = d4_x_speed, n = 100)$y,
            d1_y_speed = approx(x = 1:n(), y = d1_y_speed, n = 100)$y,
            d2_y_speed = approx(x = 1:n(), y = d2_y_speed, n = 100)$y,
            d3_y_speed = approx(x = 1:n(), y = d3_y_speed, n = 100)$y,
            d4_y_speed = approx(x = 1:n(), y = d4_y_speed, n = 100)$y)

saveRDS(clean_data_all_wide_norm, "clean_data_all_wide_norm.Rds")
```

## Experimental information

```{r}
animal_group<-data.frame(animal_id=c("4873","4874","4904","4907",
                                     "4910","4911","4913","4915","4918","4875",
                                     "4902","4903","4906","4909",
                                     "4908","4912","4914","4916","4917"),
                         group = c(rep("NT", 9), rep("NT+ES", 10)))

sex_paw<-read.csv("data/original/sex_paw_preference.csv")
colnames(sex_paw)[1]<-"animal_id"

animal_group<-sex_paw%>%
  mutate(animal_id = as.character(animal_id))%>%
  full_join(animal_group)

table(animal_group$group, animal_group$Sex)
table(animal_group$group, animal_group$paw_preference)
```


## KDI normalized

### Univariate

```{r}
all_data_df<-all_data_df%>%
  mutate(animal_id = as.character(animal_id))%>%
  left_join(animal_group)%>%
    mutate(timepoint = factor(timepoint, levels = c("pre_injury", "post-SCI", "8w-SCI", "16w-SCI")),
         reach_type = factor(reach_type, levels = c("successful", "unsuccessful", "attempt_pellet", "attempt_Nopellet")))


all_data_df%>%
  filter(timepoint == "pre_injury")%>%
ggplot(aes(frame, PC1, color = group, fill = group))+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(timepoint~reach_type)

all_data_df%>%
  filter(timepoint == "pre_injury")%>%
ggplot(aes(frame, PC2, color = group, fill = group))+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(.~reach_type)

all_data_df%>%
  filter(timepoint == "pre_injury")%>%
ggplot(aes(frame, PC3, color = group, fill = group))+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(.~reach_type)

all_data_df%>%
  filter(timepoint == "pre_injury")%>%
ggplot(aes(frame, angle, color = group, fill = group))+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(.~reach_type)

all_data_df%>%
  filter(timepoint == "pre_injury")%>%
ggplot(aes(frame, smooth_x_d1, color = group, fill = group))+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(.~reach_type)

all_data_df%>%
  filter(timepoint == "pre_injury")%>%
ggplot(aes(frame, smooth_x_d2, color = group, fill = group))+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(.~reach_type)

all_data_df%>%
  filter(timepoint == "pre_injury")%>%
ggplot(aes(frame, smooth_y_d1, color = group, fill = group))+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(.~reach_type)

all_data_df%>%
  filter(timepoint == "pre_injury")%>%
ggplot(aes(frame, smooth_y_d2, color = group, fill = group))+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(.~reach_type)

all_data_df%>%
ggplot(aes(frame, smooth_x_d1, color = paw_preference))+
  # geom_path(aes(group = reach_id), alpha = 0.1)+
  stat_summary(fun = mean, geom = "line")+
  # stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(timepoint~reach_type)


uni_raw_df<-raw_data_all%>%
  # filter(timepoint == "pre_injury")%>%
  mutate(animal_id = as.character(animal_id),
         reach_id = paste(timepoint, animal_id, reach, sep = "_"))%>%
  left_join(animal_group)

uni_raw_df%>%
ggplot(aes(x_d1, y_d1, color = group, fill = group))+
  geom_path(aes(group = reach_id), alpha = 0.3)+
  facet_grid(group~reach_type)

uni_raw_df%>%
ggplot(aes(frame, x_d1, color = paw_preference))+
  # geom_path(aes(group = reach_id), alpha = 0.3)+
  stat_summary(fun = mean, geom = "line")+
  # stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(timepoint~group)+
  xlim(0,70)
```

## Correcting for X drift

Determining the average X axis drift

```{r}
L_R_summary<-all_data_df%>%
  filter(timepoint == "pre_injury")%>%
  group_by(paw_preference, frame, reach_type)%>%
  summarise(
    m_x_d1 = mean(smooth_x_d1, na.rm = T),
    m_x_d2 = mean(smooth_x_d2, na.rm = T),
    m_x_d3 = mean(smooth_x_d3, na.rm = T),
    m_x_d4 = mean(smooth_x_d4, na.rm = T))%>%
  group_by(frame, reach_type)%>%
  mutate(
    d_x_d1 = diff(m_x_d1),
    d_x_d2 = diff(m_x_d2),
    d_x_d3 = diff(m_x_d3),
    d_x_d4 = diff(m_x_d4))%>%
  group_by(reach_type)%>%
  summarise(
    m_x_d1 = mean(d_x_d1),
    m_x_d2 = mean(d_x_d2),
    m_x_d3 = mean(d_x_d3),
    m_x_d4 = mean(d_x_d4))

mean_shift<-sapply(L_R_summary[,-1], mean)

```

### Clean data corrected

Correcting the clean data by the average drift in X axis
```{r}
clean_data_all_wide_norm_drift_R<-clean_data_all_wide_norm%>%
  mutate(animal_id = as.character(animal_id))%>%
  full_join(animal_group)%>%
  filter(paw_preference == "R")%>%
  mutate(
    smooth_x_d1 = smooth_x_d1-mean_shift[1],
    smooth_x_d2 = smooth_x_d2-mean_shift[2],
    smooth_x_d3 = smooth_x_d3-mean_shift[3],
    smooth_x_d4 = smooth_x_d4-mean_shift[4])

clean_data_all_wide_norm_drift<-clean_data_all_wide_norm%>%
  mutate(animal_id = as.character(animal_id))%>%
  full_join(animal_group)%>%
  filter(paw_preference == "L")%>%
  bind_rows(clean_data_all_wide_norm_drift_R)

clean_data_all_wide_norm_drift%>%
ggplot(aes(frame, smooth_x_d1, color = paw_preference))+
  # geom_path(aes(group = reach_id), alpha = 0.1)+
  stat_summary(fun = mean, geom = "line")+
  # stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.3)+
  facet_grid(timepoint~reach_type)
```

### KDI
#### PCA per reach

We are going to perform a PCA for each one of the reaches individually, then deal with the indeterminacies later. The advantage of this is that we do not need to time normalize.

**workflow:**

* split the clean data by reach-animal
* perform PCA on each with scaling and centering
* GPA on all loadings to deal with indeterminacy

```{r}
pre_injury_kdi_df<-clean_data_all_wide_norm_drift%>%
    filter(reach_type == "successful", 
           timepoint == "pre_injury")

pre_injury_list<-split(pre_injury_kdi_df, pre_injury_kdi_df$reach_id)

per_reach_pca<-lapply(pre_injury_list, function(x){
  prcomp(x[,6:13], scale. = T, center = T)
})

per_reach_loadings<-lapply(per_reach_pca, function(x){
  x$rotation[,1:3]
})

animal_id_vec<-unlist(lapply(pre_injury_list, function(x){
  unique(x$animal_id)
}))

## rotate the loading with GPA per animal
gpa_ref_animal<-list()
for (a in unique(animal_id_vec)){
  
  ## Extract consensus
  animal_ref_index<-which(animal_id_vec == a)
  temp_ref_loadings<-per_reach_loadings[animal_ref_index]
  gpa_ref_animal[[as.character(a)]]<-GPA(temp_ref_loadings)
}

## Rotate scores of all data
all_data_list<-split(clean_data_all_wide_norm_drift, clean_data_all_wide_norm_drift$reach_id)

animal_id_vec<-unlist(lapply(all_data_list, function(x){
  unique(x$animal_id)
}))

rotated_scores<-list()
for(i in 1:length(all_data_list)){
  ## Rotate scores
  consensus<-gpa_ref_animal[which(names(gpa_ref_animal)==
                                    animal_id_vec[i])][[1]]$consensus
  
  temp_df<-all_data_list[[i]]
  pca_data<-scale(as.matrix(temp_df[,6:13]))
  scores<-as.data.frame(pca_data%*%consensus)
  temp_df[,c("PC1", "PC2", "PC3")]<-scores[,1:3]
  rotated_scores[[i]]<-temp_df
}

all_data_df<-bind_rows(rotated_scores)
```

#### derive consensus score
```{r}
GPA_scores_ref<-all_data_df%>%
  filter(reach_type == "successful", timepoint == "pre_injury")

GPA_scores_ref_list<-split.data.frame(GPA_scores_ref, GPA_scores_ref$animal_id)

## rotate the loading with GPA per animal
gpa_ref_score_animal<-list()
for (a in names(GPA_scores_ref_list)){
  ## Extract score consensus
  temp_reach_list<-split.data.frame(GPA_scores_ref_list[[a]],
                                    GPA_scores_ref_list[[a]]$reach_id)
  temp_list<-lapply(temp_reach_list, function(x) x[,c("PC1", "PC2", "PC3")])
  res_df<-as.data.frame(GPA(temp_list)$consensus)
  colnames(res_df)<-c("cPC1", "cPC2", "cPC3")
  res_df$frame<-1:100
  res_df$animal_id<-a
  gpa_ref_score_animal[[as.character(a)]]<-res_df
}

GPA_all_scores_ref<-bind_rows(gpa_ref_score_animal)
```


#### GPA distance

```{r}
GPA_all_scores_dist<-all_data_df%>%
  full_join(GPA_all_scores_ref)%>%
  group_by(frame, reach_id)%>%
  mutate(eudist = dist(rbind(c(PC1, PC2, PC3),
                             c(cPC1, cPC2, cPC3))))

saveRDS(GPA_all_scores_dist,"GPA_all_scores_dist_drift.Rds")
```


```{r }
GPA_all_scores_dist<-readRDS("GPA_all_scores_dist_drift.Rds")

GPA_all_scores_dist_summary<-GPA_all_scores_dist%>%
  group_by(reach_id)%>%
  summarise(kdi = sum(eudist),
            timepoint = unique(timepoint),
            reach_type = unique(reach_type),
            animal_id = as.character(unique(animal_id)))%>%
  full_join(animal_group)%>%
  mutate(timepoint = factor(timepoint, levels = c("pre_injury", "post-SCI", "8w-SCI", "16w-SCI")),
         reach_type = factor(reach_type, c("successful", "unsuccessful", 
                                           "attempt_Nopellet", "attempt_pellet")))

```

# Cohort 2: GTACR
## Read data

```{r}
sheets<-readxl::excel_sheets("data/original/new_analyzed GTACR videos.xlsx")

read_sheets<-function(sheet){
  temp<-read_xlsx("data/original/new_analyzed GTACR videos.xlsx", sheet = sheet)
  temp$reach_type<-sheet
  
  return(temp)
}

gtacr_raw_df<-map(sheets,.f = read_sheets)%>%
  bind_rows()

colnames(gtacr_raw_df)<-c("frame", "x_d1", "x_d2", "x_d3", "x_d4", "y_d1", "y_d2", 
                       "y_d3", "y_d4", "x_p", "y_p","digit_spread", "pellet_d1",
                       "pellet_d2", "pellet_d3", "pellet_d4",
                       "angle", "animal_id","timepoint","reach_type")
```

## Clean all data

```{r}
## function to add the reach id per session
add_reach_id<-function(frame){
  
  reach<-vector("numeric", length = length(frame))
  count_reaches<-0
  # animal<-unique(temp$animal_id)[1]
  for (i in 1:length(frame)){
    if (frame[i] == 1){
      count_reaches<-count_reaches+1
    }
    reach[i]<-count_reaches
  }
  return(reach)
}


gtacr_raw_df<-gtacr_raw_df%>%
  arrange(animal_id)%>%
  group_by(timepoint, animal_id)%>%
  mutate(reach = add_reach_id(frame))

```

```{r, eval = FALSE}
# create unique reach identifier
gtacr_raw_df<-gtacr_raw_df%>%
  mutate(reach_id = paste(timepoint, animal_id, reach, sep = "_"))

# table(gtacr_raw_df$reach_id, gtacr_raw_df$reach_type)

## Calculate distances between digits
gtacr_raw_df<-gtacr_raw_df%>%
  group_by(reach_id, frame)%>%
  mutate(d1_d2 = dist(rbind(c(x_d1, y_d1),
                            c(x_d2, y_d2))),
         d2_d3 = dist(rbind(c(x_d2, y_d2),
                            c(x_d3, y_d3))),
         d3_d4 = dist(rbind(c(x_d3, y_d3),
                            c(x_d4, y_d4))))

# save dataset for future use
saveRDS(gtacr_raw_df, "new_gtacr_raw_df.Rda")
```

```{r}
#load raw all dataset
gtacr_raw_df<-readRDS("new_gtacr_raw_df.Rda")

#determine quantile threshods
thr_d1_d2<-quantile(gtacr_raw_df$d1_d2, probs = 0.92)
thr_d2_d3<-quantile(gtacr_raw_df$d2_d3, probs = 0.92)
thr_d3_d4<-quantile(gtacr_raw_df$d3_d4, probs = 0.92)

#Eliminate extremes filtered based on the distance between digits
gtacr_clean_df<-gtacr_raw_df%>%
  ungroup()%>%
  rowwise()%>%
  mutate(x_d1 = ifelse(d1_d2>thr_d1_d2, NA, x_d1),
         y_d1 = ifelse(d1_d2>thr_d1_d2, NA, y_d1),
         x_d2 = ifelse(d2_d3>thr_d2_d3 | d1_d2>thr_d1_d2, NA, x_d2),
         y_d2 = ifelse(d2_d3>thr_d2_d3 | d1_d2>thr_d1_d2, NA, y_d2),
         x_d3 = ifelse(d2_d3>thr_d2_d3 | d3_d4>thr_d3_d4, NA, x_d3),
         y_d3 = ifelse(d2_d3>thr_d2_d3 | d3_d4>thr_d3_d4, NA, y_d3),
         x_d4 = ifelse(d3_d4>thr_d3_d4, NA, x_d4),
         y_d4 = ifelse(d3_d4>thr_d3_d4, NA, y_d4))

# Calculate mahalanobis distance
out<-psych::outlier(gtacr_clean_df[,c(2,9)])

gtacr_clean_df<-gtacr_clean_df%>%
  ungroup()%>%
  mutate(out = out,
         filter_out = out>20)%>%## filter based on mahalanobis distance
  filter(filter_out == FALSE)
```


```{r}
# table(gtacr_clean_df$reach_id, gtacr_clean_df$reach_type)
# Summarise based on proportion of NA values. Those reaches with high NA (more tha 20%) are excluded
gtacr_clean_summary<-gtacr_clean_df%>%
  group_by(reach_id)%>%
  summarise_at(.vars = c("x_d1", "y_d1", "x_d2", "y_d2", "x_d3", "y_d3", "x_d4", "y_d4"),
               .funs = function(x) {mean(is.na(x))})

gtacr_clean_summary$means_nas<-rowMeans(gtacr_clean_summary[,-1])
reach_to_keep<-gtacr_clean_summary$reach_id[which(gtacr_clean_summary$means_nas<=0.2)]

gtacr_clean_df<-gtacr_clean_df%>%
  filter(reach_id%in%reach_to_keep)

# Transform to long format
gtacr_clean_df_long<-gtacr_clean_df%>%
  pivot_longer(cols = c(x_d1, x_d2, x_d3, x_d4, y_d1, y_d2, y_d3, y_d4, 
                        pellet_d1, pellet_d2, pellet_d3, pellet_d4))%>%
  mutate(digit=str_split(name, "_", simplify = T)[,2], 
         pos_var = str_split(name, "_", simplify = T)[,1],
         pos_var = ifelse(pos_var == "pellet", "dist_pellet", pos_var))%>%
  select(-name)%>%
  pivot_wider(id_cols = c(animal_id, timepoint, reach_id, frame,digit, x_p, y_p,
                          digit_spread, angle, reach_type), names_from = pos_var)


## Use tsclean and supsmu to clean temporal outliers and smooth
gtacr_clean_df_s<-gtacr_clean_df_long%>%
  # filter(reach_id == "After SCI_5912_1", digit =="d1")
  group_by(reach_id, digit)%>%
  mutate(n_frame= n())%>%
  filter(n_frame>=20)%>%
  filter(frame<=(n_frame-5))%>%
  mutate(clean_x = tsclean(x),
         clean_y = tsclean(y),
         clean_x_p = tsclean(x_p),
         clean_y_p = tsclean(y_p))%>%
  mutate(smooth_x = supsmu(x = frame, y = clean_x, span = 0.05)$y,
         smooth_y = supsmu(x = frame, y = clean_y, span = 0.05)$y,
         smooth_x_p = supsmu(x = frame, y = clean_x_p, span = 0.05)$y,
         smooth_y_p = supsmu(x = frame, y = clean_y_p, span = 0.05)$y)

## Add features
gtacr_clean_df_wide<-gtacr_clean_df_s%>%
  pivot_wider(id_cols = c(animal_id, timepoint, reach_id, frame, reach_type),
              names_from = "digit",
                values_from = c("smooth_x","smooth_y", "x", "y"))

gtacr_clean_df_wide<-gtacr_clean_df_wide%>%
  group_by(reach_id, frame)%>%
  mutate(d1_d4 = as.numeric(dist(rbind(c(smooth_x_d1, smooth_y_d1),
                            c(smooth_x_d4, smooth_y_d4)))),
         angle = sin((smooth_x_d1-smooth_x_d4)/d1_d4)*57.29)%>%
  group_by(reach_id)%>%
  mutate(d1_x_speed = abs(c(0,diff(smooth_x_d1))),
         d1_y_speed = abs(c(0,diff(smooth_y_d1))),
         d2_x_speed = abs(c(0,diff(smooth_x_d2))),
         d2_y_speed = abs(c(0,diff(smooth_y_d2))),
         d3_x_speed = abs(c(0,diff(smooth_x_d3))),
         d3_y_speed = abs(c(0,diff(smooth_y_d3))),
         d4_x_speed = abs(c(0,diff(smooth_x_d4))),
         d4_y_speed = abs(c(0,diff(smooth_y_d4))))
         

gtacr_clean_df_wide%>%
    filter(reach_id == "Light On_5912_4")%>%
  ggplot()+
  geom_segment(aes(x = smooth_x_d1, xend = smooth_x_d4, 
                   y = smooth_y_d1, yend = smooth_y_d4), color = "grey")+
    geom_point(aes(smooth_x_d1, smooth_y_d1))+
  geom_point(aes(smooth_x_d4, smooth_y_d4), color = "red")


gtacr_clean_df_wide%>%
    filter(animal_id == 5912, reach_type == "Successful")%>%
  ggplot()+
  geom_line(aes(frame, d4_y_speed))+
  facet_wrap(~reach_id)

saveRDS(gtacr_clean_df_wide, "new_gtacr_clean_df_wide.Rds")
```

### Check cleaning

```{r}
gtacr_raw_df%>%
    filter(timepoint == "Baseline", reach_type == "Successful",
         animal_id == "5912")%>%
  ggplot(aes(x_d3,y_d3))+
  geom_path()+
  geom_path(data = gtacr_clean_df_wide%>%
    filter(timepoint == "Baseline", reach_type == "Successful",
         animal_id == "5912"), aes(x_d3, y_d3), color = "red")+
  facet_wrap(~reach_id)

gtacr_raw_df%>%
    filter(timepoint == "Baseline", reach_type == "Successful",
         animal_id == "5912")%>%
  ggplot(aes(frame,y_d3))+
  geom_path()+
  geom_path(data = gtacr_clean_df_wide%>%
    filter(timepoint == "Baseline", reach_type == "Successful",
         animal_id == "5912"), aes(frame, smooth_y_d3), color = "red")+
  facet_wrap(~reach_id)

```

### time normalization
```{r}
gtacr_clean_df_wide_norm<-gtacr_clean_df_wide%>%
  group_by(reach_id)%>%
  reframe(animal_id = unique(animal_id),
            frame = 1:100,
            timepoint = unique(timepoint),
            reach_type = unique(reach_type),
            smooth_x_d1 = approx(x = 1:n(), y = smooth_x_d1, n = 100)$y,
            smooth_x_d2 = approx(x = 1:n(), y = smooth_x_d2, n = 100)$y, 
            smooth_x_d3 = approx(x = 1:n(), y = smooth_x_d3, n = 100)$y, 
            smooth_x_d4 = approx(x = 1:n(), y = smooth_x_d4, n = 100)$y, 
            smooth_y_d1 = approx(x = 1:n(), y = smooth_y_d1, n = 100)$y, 
            smooth_y_d2 = approx(x = 1:n(), y = smooth_y_d2, n = 100)$y, 
            smooth_y_d3 = approx(x = 1:n(), y = smooth_y_d3, n = 100)$y, 
            smooth_y_d4 = approx(x = 1:n(), y = smooth_y_d4, n = 100)$y,
            d1_d4 = approx(x = 1:n(), y = d1_d4, n = 100)$y,
            angle = approx(x = 1:n(), y = angle, n = 100)$y,
            d1_x_speed = approx(x = 1:n(), y = d1_x_speed, n = 100)$y,
            d2_x_speed = approx(x = 1:n(), y = d2_x_speed, n = 100)$y,
            d3_x_speed = approx(x = 1:n(), y = d3_x_speed, n = 100)$y,
            d4_x_speed = approx(x = 1:n(), y = d4_x_speed, n = 100)$y,
            d1_y_speed = approx(x = 1:n(), y = d1_y_speed, n = 100)$y,
            d2_y_speed = approx(x = 1:n(), y = d2_y_speed, n = 100)$y,
            d3_y_speed = approx(x = 1:n(), y = d3_y_speed, n = 100)$y,
            d4_y_speed = approx(x = 1:n(), y = d4_y_speed, n = 100)$y)

saveRDS(gtacr_clean_df_wide_norm, "new_gtacr_clean_df_wide_norm.Rds")
```

### Filtering trials

There are several trials for which the processing fails due to high inaccuracies from the markerless prediction. We will manually filter those for the purpose of understanding the KDI metrics better.

```{r}
# function to facilitate iterate through all trials to select them

raw_clean_plot<-function(a= NULL, r = NULL, tp = NULL, id = NULL){
  
  clean_plot_df<-gtacr_clean_df_wide_norm%>%
    filter(reach_id == id)%>%
    select(-reach_type, -timepoint, -animal_id)%>%
    pivot_longer(-c(reach_id, frame))%>%
    filter(str_detect(name, "smooth"))%>%
    mutate(digit = str_extract(name, "d[[:digit:]]"),
           coor = str_extract(name, "x|y"))%>%
    pivot_wider(id_cols = c(reach_id, frame, digit), names_from = 
                  "coor")
  
  raw_plot_df<-gtacr_raw_df%>%
    filter(reach_id == id)%>%
    select(-reach_type, -timepoint, -animal_id, -d1_d2, -d2_d3, -d3_d4)%>%
    pivot_longer(-c(reach_id, frame))%>%
    filter(str_detect(name, "_d"))%>%
    mutate(digit = str_extract(name, "d[[:digit:]]"),
           coor = str_extract(name, "x|y"))%>%
    pivot_wider(id_cols = c(reach_id, frame, digit), names_from = 
                  "coor")
    
  ggplot(clean_plot_df,aes(x, y, group = digit, color = digit))+
    geom_path()+
    geom_point(data = clean_plot_df%>%filter(frame == 1))+
    geom_point(data = clean_plot_df%>%filter(frame == 100), shape = 21)+
    facet_wrap(~reach_id)+
    theme_minimal()+
    xlab("x coordinate")+ylab("y coordinate")+
    theme(legend.position = c(0.85, 0.12))+
    coord_equal()
}

# # Iteration through all the trials
# iterator_df<-expand.grid(a = unique(gtacr_clean_df_wide_norm$animal_id),
#                          r = unique(gtacr_clean_df_wide_norm$reach_type),
#                          tp = unique(gtacr_clean_df_wide_norm$timepoint))
# 
# reach_ids<-unique(gtacr_clean_df_wide_norm$reach_id)
# res<-vector(length = length(reach_ids))
# for(i in 1:length(reach_ids)){
#   plot(raw_clean_plot(id = reach_ids[i]))
#   res[i]<-readline("Y/N")
# }
# 
# # logical vector containing whether to keep (T) or not (F) a specific trial
# res[res=="FF"]<-F
# res<-as.logical(res)
# 
# reach_keep<-reach_ids[res]
# saveRDS(res, "res_filter_reach_id.Rds")
# write.csv(reach_ids[!res], "res_filter_reach_id.csv", row.names = F)
```

```{r}
# selected trial data
gtacr_clean_df_wide_norm_f<-gtacr_clean_df_wide_norm
```

## KDI normalized
### PCA per reach 

We are going to perform a PCA for each one of the reaches individually, then deal with the indeterminacy later. The advantage of this is that we do not need to time normalize.

**workflow:**

* split the clean data by reach-animal
* perform PCA on each with scaling and centering
* GPA on all loadings to deal with indeterminacy

**PCA per reach**

```{r}
# clean_data_all_wide_norm<-readRDS("clean_data_all_wide_norm.Rds")

baseline_kdi_df<-gtacr_clean_df_wide_norm_f%>%
    filter(reach_type == "Successful", 
           timepoint == "Baseline")

gtacr_clean_df_wide_norm_f<-gtacr_clean_df_wide_norm_f%>%
  filter(animal_id %in% baseline_kdi_df$animal_id)

baseline_list<-split(baseline_kdi_df, baseline_kdi_df$reach_id)

bl_reach_pca<-lapply(baseline_list, function(x){
  prcomp(x[,6:13], scale. = T, center = T)
})

bl_reach_loadings<-lapply(bl_reach_pca, function(x){
  x$rotation[,1:3]
})

animal_id_vec<-unlist(lapply(baseline_list, function(x){
  unique(x$animal_id)
}))

## rotate the loading with GPA per animal
gpa_ref_animal<-list()
for (a in unique(animal_id_vec)){
  
  ## Extract consensus
  animal_ref_index<-which(animal_id_vec == a)
  temp_ref_loadings<-bl_reach_loadings[animal_ref_index]
  if (length(temp_ref_loadings)<=1){
    gpa_ref_animal[[as.character(a)]]<-list(consensus=as.data.frame(temp_ref_loadings))
  }else{
      gpa_ref_animal[[as.character(a)]]<-GPA(temp_ref_loadings)
  }
}

## Rotate scores of all data
gtacr_all_data_list<-split(gtacr_clean_df_wide_norm_f, gtacr_clean_df_wide_norm_f$reach_id)

animal_id_vec<-unlist(lapply(gtacr_all_data_list, function(x){
  unique(x$animal_id)
}))

gtacr_rotated_scores<-list()
for(i in 1:length(gtacr_all_data_list)){
  ## Rotate scores
  consensus<-gpa_ref_animal[which(names(gpa_ref_animal)==
                                    animal_id_vec[i])][[1]]$consensus
  colnames(consensus)<-c("PC1", "PC2", "PC3")
  
  temp_df<-gtacr_all_data_list[[i]]
  pca_data<-scale(as.matrix(temp_df[,6:13]))
  scores<-as.data.frame(pca_data%*%as.matrix(consensus))
  temp_df[,c("PC1", "PC2", "PC3")]<-scores[,1:3]
  gtacr_rotated_scores[[i]]<-temp_df
}

gtacr_all_data_df<-bind_rows(gtacr_rotated_scores)
saveRDS(gtacr_all_data_df, "gtacr_all_data_df.Rds")
```

#### derive consensus score
```{r}
gtacr_GPA_scores_ref<-gtacr_all_data_df%>%
      filter(reach_type == "Successful", 
           timepoint == "Baseline")

gtacr_GPA_scores_ref_list<-split.data.frame(gtacr_GPA_scores_ref, gtacr_GPA_scores_ref$animal_id)

## rotate the loading with GPA per animal
gtacr_gpa_ref_score_animal<-list()
for (a in names(gtacr_GPA_scores_ref_list)){
  ## Extract score consensus
  temp_reach_list<-split.data.frame(gtacr_GPA_scores_ref_list[[a]],
                                    gtacr_GPA_scores_ref_list[[a]]$reach_id)
  temp_list<-lapply(temp_reach_list, function(x) x[,c("PC1", "PC2", "PC3")])
  
  if (length(temp_list)<=1){
    res_df<-as.data.frame(temp_list)
  }else{
    res_df<-as.data.frame(GPA(temp_list)$consensus)
  }
  
  colnames(res_df)<-c("cPC1", "cPC2", "cPC3")
  res_df$frame<-1:100
  res_df$animal_id<-a
  gtacr_gpa_ref_score_animal[[as.character(a)]]<-res_df
}

gtacr_GPA_all_scores_ref<-bind_rows(gtacr_gpa_ref_score_animal)
```


```{r}
gtacr_GPA_all_scores_dist<-gtacr_all_data_df%>%
  mutate(animal_id=as.character(animal_id))%>%
  full_join(gtacr_GPA_all_scores_ref)%>%
  group_by(frame, reach_id)%>%
  mutate(eudist = dist(rbind(c(PC1, PC2, PC3),
                             c(cPC1, cPC2, cPC3))))

saveRDS(gtacr_GPA_all_scores_dist,"gtacr_GPA_all_scores_dist.Rds")

gtacr_GPA_all_scores_dist_summary<-gtacr_GPA_all_scores_dist%>%
  group_by(reach_id)%>%
  summarise(kdi = sum(eudist),
            timepoint = unique(timepoint),
            reach_type = unique(reach_type),
            animal_id = as.character(unique(animal_id)))%>%
  mutate(reach_type = case_when(
    reach_type == "Miss"~"Unsuccessful",
    TRUE ~ reach_type
  ))%>%
  mutate(reach_type = factor(reach_type, c("Successful", "Unsuccessful", 
                                           #"Attempts without Pellet",
                                           "Attempts with pellet")))
```

# Analysis and Figures
## Pre injury analysis

### Cohort 1: NT study
```{r}
## KDI fit for pre injury reach types
nt_fit_pre_injury<-lmer(kdi~reach_type+(1|animal_id), 
                     data = GPA_all_scores_dist_summary%>%
                       filter(timepoint=="pre_injury", reach_type!="attempt_Nopellet"))

anova(nt_fit_pre_injury)

sjPlot::plot_model(nt_fit_pre_injury, type = "pre", terms = c("reach_type"))
emmeans::emmeans(nt_fit_pre_injury, pairwise~reach_type)
```

### Cohort 2: GTACR study

```{r}
## KDI fit for pre injury reach types
gtacr_fit_pre_injury<-lmer(kdi~reach_type+(1|animal_id), 
                     data = gtacr_GPA_all_scores_dist_summary%>%
                       filter(timepoint=="Baseline", reach_type!="Attempts without Pellet"))

anova(gtacr_fit_pre_injury)

sjPlot::plot_model(gtacr_fit_pre_injury, type = "pre", terms = c("reach_type"))
emmeans::emmeans(gtacr_fit_pre_injury, pairwise~reach_type)
```

## Post Injury analysis
### Cohort 1: NT study
```{r }
# post SCI injury
nt_fit_post<-lmer(kdi~timepoint + (timepoint|animal_id), data = GPA_all_scores_dist_summary%>%
                    filter(reach_type!="attempt_Nopellet", timepoint %in% c("pre_injury", "post-SCI")))

anova(nt_fit_post)

sjPlot::plot_model(nt_fit_post, type = "pre", terms = c("timepoint"))
emmeans::emmeans(nt_fit_post, pairwise~timepoint)

# Fit by reachtype
nt_fit_byreach<-lmer(kdi~timepoint*reach_type + (timepoint|animal_id), data = 
                       GPA_all_scores_dist_summary%>%filter(reach_type!="attempt_Nopellet", 
                                                         timepoint %in% c("pre_injury", "post-SCI")))

anova(nt_fit_byreach)

sjPlot::plot_model(nt_fit_byreach, type = "pre", terms = c("timepoint", "reach_type"))
emmeans::emmeans(nt_fit_byreach, pairwise~reach_type|timepoint)

## fit KD trajectories pre injury
nt_fit_eudits<-lmer(eudist~ns(frame, 7)*factor(reach_type)+(frame|reach_id), 
              data = GPA_all_scores_dist%>%
                filter(timepoint == "pre_injury", animal_id == "4910"))

anova(fit_eudits_4910)
sjPlot::plot_model(nt_fit_eudits, type = "pre", terms = c("frame [all]","reach_type"))

GPA_all_scores_dist%>%
  filter(timepoint == "pre_injury", animal_id == "4910")%>%
ggplot(aes(frame, eudist, group = reach_id, color = reach_type))+
  # geom_line()+
  stat_summary(fun.data = mean_se, geom = "ribbon", 
               aes(group = reach_type, fill = reach_type), alpha = 0.3)+
  stat_summary(fun = "mean", geom = "line", aes(group = reach_type))

```

### Cohort 2: GTACR study
```{r }
gtacr_GPA_all_scores_dist_summary<-gtacr_GPA_all_scores_dist_summary%>%
  mutate(timepoint = factor(timepoint, 
                            levels = c("Baseline","LightOn","AfterSCI",
                                       "BaselineAfterSCI",
                                       "LightOnAfterRehab")))

# post SCI injury
gtacr_fit_post<-lmer(kdi~timepoint + (timepoint|animal_id), 
                     data = gtacr_GPA_all_scores_dist_summary%>%
                       filter(!timepoint %in% c("BaselineAfterSCI", 
                                                "LightOnAfterRehab"),
                              reach_type!="Attempts without Pellet"))

anova(gtacr_fit_post)
sjPlot::plot_model(gtacr_fit_post, type = "pre", terms = c("timepoint"))
emmeans::emmeans(gtacr_fit_post, pairwise~timepoint)

# Fit by reachtype
gtacr_fit_byreach<-lmer(kdi~timepoint*reach_type + (timepoint|animal_id), data = gtacr_GPA_all_scores_dist_summary%>%
                       filter(
                          !timepoint %in% c("BaselineAfterSCI", 
                                            "LightOnAfterRehab"),
                              reach_type!="Attempts without Pellet"))

anova(gtacr_fit_byreach)

sjPlot::plot_model(gtacr_fit_byreach, type = "pre", terms = c("timepoint", "reach_type"))

emmeans::emmeans(gtacr_fit_byreach, pairwise~reach_type|timepoint)
```

## Plots and Figures
### Plot examples

```{r }
## Selected plots
clean_plot_df<-clean_data_all_wide_norm_drift%>%
  filter(animal_id=="4915", reach_type == "successful",timepoint == "pre_injury")%>%
  select(-reach_type, -timepoint, -animal_id, -paw_preference, -Sex, -group)%>%
  pivot_longer(-c(reach_id, frame))%>%
  filter(str_detect(name, "smooth"))%>%
  mutate(digit = str_extract(name, "d[[:digit:]]"),
         coor = str_extract(name, "x|y"))%>%
  pivot_wider(id_cols = c(reach_id, frame, digit), names_from = 
                "coor")
### Plot XY
clean_plot_df%>%
  filter(reach_id %in% c("pre_injury_4915_18", "pre_injury_4915_20"))%>%
ggplot(aes(x, y, group = digit, color = digit))+
  geom_path()+
  geom_point(data = clean_plot_df%>%filter(frame == 1,
                                           reach_id %in% c("pre_injury_4915_18", "pre_injury_4915_20")))+
  geom_point(data = clean_plot_df%>%filter(frame == 100,
                                           reach_id %in% c("pre_injury_4915_18", "pre_injury_4915_20")), shape = 21)+
  facet_wrap(~reach_id)+
  theme_minimal()+
  xlab("x coordinate")+ylab("y coordinate")+
  theme(legend.position = c(0.85, 0.12))+
  coord_equal()

```


```{r}
GPA_all_scores_dist<-GPA_all_scores_dist%>%
  mutate(animal_id = as.character(animal_id))%>%
  full_join(animal_group)%>%
  mutate(timepoint = factor(timepoint, levels = c("pre_injury", "post-SCI", "8w-SCI",  "16w-SCI")),
         eudist = as.numeric(eudist),
         reach_type = factor(reach_type, levels = c("successful", "unsuccessful", "attempt_pellet", "attempt_Nopellet")))
```

### Figure 3

```{r}

## prediction fit_pre_injury
Fig3a<-ggeffects::ggpredict(nt_fit_pre_injury)[[1]]%>%
  ggplot(aes(x, predicted))+
  geom_point(size =2, col = "red")+
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.3, col = "red")+
  theme_minimal()+
  ylim(80, 270)+
  xlab(NULL)+ylab("KDI (estimated  95% CI)")+
  scale_x_discrete(labels = c("Success", "Grasping fail", "Reaching fail"))

Fig3b<-ggeffects::ggpredict(gtacr_fit_pre_injury)[[1]]%>%
  ggplot(aes(x, predicted))+
  geom_point(size =2, col = "red")+
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.3, col = "red")+
  theme_minimal()+
  ylim(80, 270)+
  xlab(NULL)+ylab("KDI (estimated  95% CI)")+
  scale_x_discrete(labels = c("Success", "Grasping fail", "Reaching fail"))

#Fig 3c
## fit KD trajectories pre injury for a single animal example 4910
fit_eudits_4910<-lmer(eudist~ns(frame, 7)*factor(reach_type)+(frame|reach_id), 
              data = GPA_all_scores_dist%>%
                filter(timepoint == "pre_injury", animal_id == 4910, reach_type!="attempt_Nopellet"))
pred_eudits_4910<-ggeffects::ggpredict(fit_eudits_4910, terms = c("frame [all]", "reach_type"))

pred_eudits_4910<-pred_eudits_4910%>%
  as.data.frame()%>%
  mutate(group = case_when(
    group == "unsuccessful"~"Grasping fail",
    group == "successful"~"Success",
    group == "attempt_pellet"~"Reaching fail"))
fig3c<-ggplot(pred_eudits_4910,aes(x, predicted, group = group))+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.3)+
  geom_line(aes(color = group))+
  theme_minimal()+
  xlab("% time of trial")+ylab("Distance to reference\n(estimated  95% CI)")+
  theme(legend.position = c(0.3,0.85), legend.title = element_blank())


fig3<-(Fig3a+Fig3b)/fig3c+plot_annotation(tag_levels = "a")
ggsave("plots/kdi_fig3.png", fig3, width = 8, height = 6)
```

### Figure 4

```{r}

fig4a<-ggeffects::ggpredict(nt_fit_post)[[1]]%>%
  ggplot(aes(x, predicted))+
  geom_point(size =2, col = "red")+
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.3, col = "red")+
  theme_minimal()+
  xlab(NULL)+ylab("KDI (estimated  95% CI)")+
  scale_x_discrete(labels = c("Pre SCI", "chronic SCI"))+
  ylim(100, 320)+
  ggtitle("Cohort 1")

fig4b<-ggeffects::ggpredict(nt_fit_byreach, terms = c("timepoint", "reach_type"))%>%
  as.data.frame()%>%
  mutate(group = case_when(
    group == "unsuccessful"~"Grasping fail",
    group == "successful"~"Success",
    group == "attempt_pellet"~"Reaching fail"),
    group = factor(group, levels = c("Success", "Grasping fail", "Reaching fail")))%>%
  ggplot(aes(x, predicted, color = group))+
  geom_point(size =2, position = position_dodge(width = 0.5))+
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.3,
                position = position_dodge(width = 0.5))+
  theme_minimal()+
  xlab(NULL)+ylab("KDI (estimated  95% CI)")+
  scale_x_discrete(labels = c("Pre SCI", "chronic SCI"))+
  ylim(100, 320)+
  ggtitle("Cohort 1")

fig4c<-ggeffects::ggpredict(gtacr_fit_post)[[1]]%>%
  ggplot(aes(x, predicted))+
  geom_point(size =2, col = "red")+
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.3, col = "red")+
  theme_minimal()+
  xlab(NULL)+ylab("KDI (estimated  95% CI)")+
  scale_x_discrete(labels = c("Pre SCI", "M1 silencing (Pre SCI)", "Acute SCI"))+
  ylim(100, 320)+
  ggtitle("Cohort 2")

fig4d<-ggeffects::ggpredict(gtacr_fit_byreach, terms = c("timepoint", "reach_type"))%>%
  as.data.frame()%>%
  mutate(group = case_when(
    group == "Unsuccessful"~"Grasping fail",
    group == "Successful"~"Success",
    group == "Attempts with pellet"~"Reaching fail"),
    group = factor(group, levels = c("Success", "Grasping fail", "Reaching fail")))%>%
  ggplot(aes(x, predicted, color = group))+
  geom_point(size =2, position = position_dodge(width = 0.5))+
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.3,
                position = position_dodge(width = 0.5))+
  theme_minimal()+
  xlab(NULL)+ylab("KDI (estimated  95% CI)")+
  ylim(100, 320)+
  scale_x_discrete(labels = c("Pre SCI", "M1 silencing (Pre SCI)", "Acute SCI"))+
  ggtitle("Cohort 2")


## fit KD trajectories for a single animal example

gtacr_GPA_all_scores_dist%>%
  filter(timepoint %in% c("LightOn", "Baseline"), 
         !reach_type %in% c("Attempts without Pellet", "Attempts with Pellet"), 
         animal_id == "5912")%>%
  ggplot(aes(frame, eudist, group = reach_id, color = timepoint))+
  geom_line()+
  facet_wrap(~reach_type)


fit_eudits_5912<-lmer(eudist~ns(frame, 7)*timepoint+(frame|reach_id), 
              data = gtacr_GPA_all_scores_dist%>%
                filter(timepoint %in% c("LightOn", "Baseline"), 
         reach_type =="Successful", animal_id == "5912"))

pred_eudits_5912<-ggeffects::ggpredict(fit_eudits_5912, terms = c("frame [all]", "timepoint"))

emmeans::emmeans(fit_eudits_5912, pairwise~timepoint|frame, 
                 at = list(frame = seq(1,101,10)))

fig4e<-ggplot(pred_eudits_5912,aes(x, predicted, group = group))+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.3)+
  geom_line(aes(color = group))+
  theme_minimal()+
  xlab("% time of trial")+ylab("Distance to reference\n(estimated  95% CI)")+
  theme(legend.position = c(0.2,0.85), legend.title = element_blank())+
  ylim(0,4.5)+
  ggtitle("Example animal from Cohort 2: Successful trials")

fit_eudits_5912b<-lmer(eudist~ns(frame, 7)*timepoint+(frame|reach_id), 
              data = gtacr_GPA_all_scores_dist%>%
                filter(timepoint %in% c("LightOn", "Baseline"), 
         reach_type =="Miss", animal_id == "5912"))

emmeans::emmeans(fit_eudits_5912b, pairwise~timepoint|frame, 
                 at = list(frame = seq(1,100, length = 10)))

pred_eudits_5912b<-ggeffects::ggpredict(fit_eudits_5912b, terms = c("frame [all]", "timepoint"))

fig4f<-ggplot(pred_eudits_5912b,aes(x, predicted, group = group))+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.3)+
  geom_line(aes(color = group))+
  theme_minimal()+
  xlab("% time of trial")+ylab("Distance to reference\n(estimated  95% CI)")+
  theme(legend.position = c(0.2,0.85), legend.title = element_blank())+
  ylim(0,4.5)+
  ggtitle("Example animal from Cohort 2: Grasping fail trials")

fig4_layaout<-c(
  area(1,1), area(1,2),
  area(2,1), area(2,2),
  area(3,1, r = 2),
  area(4,1, r = 2)
)

fig4<-fig4a+fig4b+fig4c+fig4d+fig4e+fig4f+plot_layout(design = fig4_layaout, guides = "collect")+plot_annotation(tag_levels = "a")

ggsave("plots/kdi_dev_fig4.png", fig4, width = 10, height = 10)
```

## Supplementary Figure

### Sub Fig 1

```{r}
GPA_all_scores_dist_summary%>%
  filter(timepoint=="pre_injury")%>%
ggplot(aes(reach_type,kdi, color = reach_type))+
  stat_summary(fun = "mean", geom = "point")+
  stat_summary(fun.data = mean_se, geom = "errorbar")+
  facet_wrap(.~animal_id, nrow = 3)

## KD trajectories for pre injury reach types
GPA_all_scores_dist%>%
  filter(timepoint == "pre_injury")%>%
ggplot(aes(frame, eudist, group = reach_id, color = animal_id))+
  # geom_line(alpha = 0.1)+
  stat_summary(fun = "mean", geom = "line", 
               aes(group = animal_id), show.legend = F)+
  facet_wrap(.~reach_type)
```


### data processing
```{r}
### Plot y
clean_plot_df_noNorm<-clean_data_all_wide%>%
  filter(animal_id=="4875", reach_type == "successful",timepoint == "pre_injury")

figS1a<-ggplot(clean_plot_df_noNorm, aes(frame, x_d1))+
  geom_line()+
  geom_point(alpha = 0.4)+
  geom_line(aes(y = smooth_x_d1), color = "red")+
  facet_wrap(~reach_id, nrow = 2)+
  theme_minimal()+
  ylab("y coordinate")

#Individual animals
clean_plot_df<-clean_data_all_wide_norm%>%
  filter(animal_id=="4875", reach_type == "successful",timepoint == "pre_injury")%>%
  select(-reach_type, -timepoint, -animal_id)%>%
  pivot_longer(-c(reach_id, frame))%>%
  filter(str_detect(name, "smooth"))%>%
  mutate(digit = str_extract(name, "d[[:digit:]]"),
         coor = str_extract(name, "x|y"))%>%
  pivot_wider(id_cols = c(reach_id, frame, digit), names_from = 
                "coor")

raw_plot_df<-raw_data_all_long%>%
  filter(animal_id=="4875", reach_type == "successful",timepoint == "pre_injury")%>%
  select(-reach_type, -timepoint, -animal_id)
  
### Plot X
figS1b<-ggplot(clean_plot_df,aes(x, y, group = digit, color = digit))+
  geom_path(show.legend = F)+
  geom_point(data = clean_plot_df%>%filter(frame == 1),show.legend = F)+
  geom_point(data = clean_plot_df%>%filter(frame == 100), shape = 21,show.legend = F)+
  facet_wrap(~reach_id, nrow = 2)+
  theme_minimal()+
  xlab("x coordinate")+ylab("y coordinate")
  # theme(legend.position = c(0.85, 0.12))+
  # coord_equal()

figS1c<-ggplot(clean_plot_df,aes(frame, y, group = reach_id, color = digit))+
  geom_path(show.legend = F)+
  facet_wrap(~digit, nrow = 1)+
  theme_minimal()+
  xlab("Time (% of task sequence)")+ylab("y coordinate")
  # theme(legend.position = c(0.85, 0.12))
  # coord_equal()
  
figS1d<-ggplot(clean_plot_df,aes(frame, x, group = reach_id, color = digit))+
  geom_path(show.legend = F)+
  facet_wrap(~digit, nrow = 1)+
  theme_minimal()+
  xlab("Time (% of task sequence)")+ylab("x coordinate")
  # theme(legend.position = c(0.85, 0.12))
  # coord_equal()

figS1<-figS1a+figS1c+figS1d+figS1b+plot_layout(ncol = 1)+plot_annotation(tag_levels = "a")
ggsave("figs1.png", figS1, height = 12, width = 7)
```


